Measuring matter effects in Neutron star - Black hole binaries with Advanced LIGO

### 2016-04-04 Paper/paper.tex
fleshing out s5

In Ref.~\cite{Lackey:2013axa}, the authors presented a suite of $134$ numerical
relativity simulations of neutron stars inspiraling into spinning black holes. Amongst
the varied quantities were (a) the equation of state for the NS, (b) NS mass, 
(c) mass-ratio and (d) black hole spins, taken as aligned with the orbit.
A total of $21$ two-parameter EoSs were used for these simulations, which also 
had $q\in[2, 5]$, $\chibh\in[-0.5, +0.75]$, and 
$m_\mathrm{NS}\in[1.20, 1.45]M_\odot$. Based on this suite of simulations, the
authors calibrated a frequency-domain waveform model for the inspiral and merger phasing of
NSBH coalescences. The GW strain predicted by this model (henceforth ``LEA''),
can be written as
% 
\begin{equation}
 \tilde{h}_\mathrm{NSBH}(f, \vec{\theta}) = \tilde{h}_\mathrm{BBH}(f, \vec{\theta} - \{\lambdans\})\,A(f, \vec{\theta})\,e^{\ii \Delta\Phi(f, \vec{\theta})},
\end{equation}
where $\vec{\theta}:=\{M, \eta, \chibh, \lambdans\}$ are the intrinsic binary
parameters. The first term on the right hand side is an underlying
BBH waveform model that is used as a base. In the original construction of LEA,
this was taken to be the SEOBNRv1 model~\cite{Taracchini:2012} of the 
Effective-one-body (EOB) family~\cite{Buonanno99}. The second term changes the 
amplitude of the BBH model
to match that of an NSBH merger of otherwise identical parameters, with NS-matter
effects parametrized by $\lambdans$. During inspiral this term is set to
unity, but is a sensitive function of $\lambdans$ close to merger. The third 
term does the same as the second, but to the waveform phasing. During inspiral,
$\Delta\Phi(f, \vec{\theta})$ is set to the PN tidal phasing corrections,
at the leading and next-to-leading orders~\cite{Vines2011}. Close to merger,
additional phenomenological terms are added and calibrated to the $134$ available
NR simulations.

@@

The monte carlo algorithm samples the parameter space in proportion with 
the likelihood $\LL(\vec{\theta}) = p(d|\vec{\theta})$, with a 
uniform prior in parameters $p(\vec{\theta}) = \Theta(\vec{\theta}-\vec{\theta}_L) + \Theta(\vec{\theta}_H-\vec{\theta})$
\begin{equation}
 p(\vec{\theta}|d) = p(d|\vec{\theta}) p(\vec{\theta})
\end{equation}

@@

\textbf{Astrophysical source population: }\label{s2:astro_multiple}
% 
Imagine we have $N$ stretches of data,
$d_1, d_2, \cdots, d_N$, each containing a single observation of an NSBH binary. Let $K$
denote our collective prior knowledge, except for expectations on binary parameters
themselves, which will enter the following calculation explicitly. Let the gravitational wave
signal from each binary be characterized by the parameters 
$\vec{\theta} = \{\mbh, \mns, \chibh, \chins, \rho\}$ and $\lambdans$. Note that we have folded
in the luminosity distance, inclination angle of the binary's orbital angular momentum with
respect to the plane of the detector, and its sky location angles into a single parameter $\rho$,
which is the SNR of the received signal. Note also that $\vec{\theta}$ does not include
$\lambdans$ as the latter would be treated independently. In addition, we assume
that all observed NS's have the same equation of state, that uniquely determines their
$\lambdans$ parameter. In other words, we assume that we know its mass perfectly.
% 
Using Bayes' theorem, the measured probability distribution function for the $\lambdans$
parameter is given as
\begin{equation}\label{eq:lambdaMultiple}
 p(\lambdans | d_1, d_2, \cdots, d_N; K) = p(\lambdans | K)^{\alpha}\prod_{i=1}^N p(\lambdans | d_i, K),
\end{equation}
where $p(\lambdans|K)$ is the prior expectation for $\lambdans$, and we have assumed
that all of the $N$ observations are independent. [Note: $\alpha$ will be calculated 
soon, it should be a function only of N.]. A priori, we will assume that no particular value
of $\lambdans$ is more preferred over another, and $\lambdans\leq 4000$, ie 
$$
p(\lambdans | K) = \mathrm{Rect}\left(\frac{\lambdans-2000}{4000}\right)
$$
% 
The second set of terms $p(\lambdans | d_i, K)$ are the inferred posterior distribution
functions for $\lambdans$ from \textit{each} observation independently, and are 
calculated by marginalizing over all other binary parameters, i.e.
\begin{equation}
 p(\lambdans | d_n, K) = \int \D\, \vec{\theta} p(\vec{\theta}, \lambdans | d_n, K).
\end{equation}
The joint inferred probability distribution of all parameters $\vec{\theta}\cup\{\lambdans\}$ 
is given by
\begin{equation}
 p(\vec{\theta}, \lambdans | d_n, K) = \dfrac{p(d_n|\vec{\theta}, \lambdans, K)\,p(\vec{\theta}, \lambdans | K)}{p(d_n|K)}.
\end{equation}
Here, $p(\vec{\theta}, \lambdans | K)$ is the prior probability for binary parameters of taking 
particular values. In this paper, we use a uniform prior on the individual masses within 
pre-chosen ranges, as well as a uniform prior on the spin of the black hole. We set neutron star's
spin identically $=0$, and its tidal deformability parameter $\lambdans$ to be equally likely
between $[0, 4000]$. The prior probability of obtaining a particular realization of data
$p(d_n|K)$ is absorbed into the overall normalization. Finally, the first term in the numerator
$p(d_n|\vec{\theta}, \lambdans, K)$ is the likelihood of obtaining the given stretch of 
data $d_n$ if we assume that a signal parameterized by $\vec{\theta}\cup\{\lambdans\}$ is buried
in it, i.e.
\begin{equation}\label{eq:likelihood}
 \LL(\vec{\theta}, \lambdans) \equiv p(\vec{\theta}, \lambdans | K) = \mathcal{N} \mathrm{exp}[- \langle d_n - h | d_n - h\rangle ],
\end{equation}
where $h\equiv h(\vec{\theta}, \lambdans)$ is our waveform model, and the detector noise
weighted inner-product $\langle a|b\rangle$ is defined as 
$$
\langle a|b\rangle \equiv 4\,\mathrm{Re}\left[\int_0^\infty \dfrac{\tilde{a}(f) \tilde(b)(f)^*}{S_n(|f|)}\,\D f\right],
$$
where $\tilde{a}$ means the Fourier transform of the finite time series $a(t)$, and 
$S_n(|f|)$ is the one-sided amplitude spectrum of detector noise. The definition of $\LL$ in 
Eq.~\ref{eq:likelihood} assumes that the instrument noise is colored Gaussian.

\prayush{%
As described in the previous sections, we have measured the $p(\lambdans | d_i, K)$ for a 
set of different binary parameters. Therefore, the problem of finding the cumulative measurement
constraint on $\lambdans$ using Eq.~\ref{eq:lambdaMultiple} reduces to being able to choose
an astrophysically motivated sample set of $N$ binaries over which the second term on the RHS 
can be computed.}

\prayush{%
Get posterior measurements for low SNRs, i.e. $\rho=\{10, 15, 20, 25, 30\}$. 
Fix black hole mass and spins. Compute the measurement certainty for $\lambdans$
as a function of the number of detections for a population astrophysically distributed 
in effective distance. Repeat for a few different combinations of black hole masses
and spins. This would be a sub-set of the overall detected population for NSBH binaries,
but what fraction of it would this comprise of is something that we can only probe with
LIGO itself. We do not know the real distribution of low-mass black holes in the field.
}

>>>>

In Ref.~\cite{Lackey:2013axa}, the authors presented a suite of $134$ numerical
relativity simulations of neutron stars inspiraling into spinning black holes. Amongst
the varied quantities were (a) the equation of state for the NS, (b) NS mass, 
(c) mass-ratio and (d) black hole spins, taken as aligned with the orbit.
A total of $21$ two-parameter EoSs were used for these simulations, which also 
had $q\in[2, 5]$, $\chibh\in[-0.5, +0.75]$, and 
$m_\mathrm{NS}\in[1.20, 1.45]M_\odot$. Based on this suite of simulations, the
authors calibrated a frequency-domain waveform model for the inspiral and merger phasing of
NSBH coalescences. The GW strain predicted by this model (henceforth ``LEA''),
can be written as
% 
\begin{equation}
 \tilde{h}_\mathrm{NSBH}(f, \vec{\theta}) = \tilde{h}_\mathrm{BBH}(f, \vec{\theta} - \{\lambdans\})\,A(f, \vec{\theta})\,e^{\ii \Delta\Phi(f, \vec{\theta})},
\end{equation}
where $\vec{\theta}:=\{\mbh, \mns, \chibh,\chins=0, \lambdans\}$ are the intrinsic binary
parameters. The first term on the right hand side is an underlying
BBH waveform model that is used as a base. In the original construction of LEA,
this was taken to be the SEOBNRv1 model~\cite{Taracchini:2012} of the 
Effective-one-body (EOB) family~\cite{Buonanno99}. The second term changes the 
amplitude of the BBH model
to match that of an NSBH merger of otherwise identical parameters, with NS-matter
effects parametrized by $\lambdans$. During inspiral this term is set to
unity, but is a sensitive function of $\lambdans$ close to merger. The third 
term does the same as the second, but to the waveform phasing. During inspiral,
$\Delta\Phi(f, \vec{\theta})$ is set to the PN tidal phasing corrections,
at the leading and next-to-leading orders~\cite{Vines2011}. Close to merger,
additional phenomenological terms are added and calibrated to the $134$ available
NR simulations.

@@

The joint inferred probability distribution of all parameters $\vec{\theta}\cup\{\lambdans\}$ 
is given by
\begin{equation}
 p(\vec{\theta}, \lambdans | d_n, K) = \dfrac{p(d_n|\vec{\theta}, \lambdans, K)\,p(\vec{\theta}, \lambdans | K)}{p(d_n|K)}.
\end{equation}
Here, $p(\vec{\theta}, \lambdans | K)$ is the prior probability for binary parameters of taking 
particular values. In this paper, we use a uniform prior on the individual masses within 
pre-chosen ranges, as well as a uniform prior on the spin of the black hole. We set neutron star's
spin identically $=0$, and its tidal deformability parameter $\lambdans$ to be equally likely
between $[0, 4000]$. The prior probability of obtaining a particular realization of data
$p(d_n|K)$ is absorbed into the overall normalization. Finally, the first term in the numerator
$p(d_n|\vec{\theta}, \lambdans, K)$ is the likelihood of obtaining the given stretch of 
data $d_n$ if we assume that a signal parameterized by $\vec{\theta}\cup\{\lambdans\}$ is buried
in it, i.e.
\begin{equation}\label{eq:likelihood}
 \LL(\vec{\theta}, \lambdans) \equiv p(\vec{\theta}, \lambdans | K) = \mathcal{N} \mathrm{exp}[- \langle d_n - h | d_n - h\rangle ],
\end{equation}
where $h\equiv h(\vec{\theta}, \lambdans)$ is our waveform model, and the detector noise
weighted inner-product $\langle a|b\rangle$ is defined as 
$$
\langle a|b\rangle \equiv 4\,\mathrm{Re}\left[\int_0^\infty \dfrac{\tilde{a}(f) \tilde(b)(f)^*}{S_n(|f|)}\,\D f\right],
$$
where $\tilde{a}$ means the Fourier transform of the finite time series $a(t)$, and 
$S_n(|f|)$ is the one-sided amplitude spectrum of detector noise. The definition of $\LL$ in 
Eq.~\ref{eq:likelihood} assumes that the instrument noise is colored Gaussian.

@@

\textbf{Astrophysical source population: }\label{s2:astro_multiple}
% 
Imagine we have $N$ stretches of data,
$d_1, d_2, \cdots, d_N$, each containing a single observation of an NSBH binary. Let $K$
denote our collective prior knowledge, except for expectations on binary parameters
themselves, which will enter the following calculation explicitly. Let the gravitational wave
signal from each binary be characterized by the parameters 
$\vec{\theta} = \{\mbh, \mns, \chibh, \chins, \rho\}$ and $\lambdans$. Note that we have folded
in the luminosity distance, inclination angle of the binary's orbital angular momentum with
respect to the plane of the detector, and its sky location angles into a single parameter $\rho$,
which is the SNR of the received signal. Note also that $\vec{\theta}$ does not include
$\lambdans$ as the latter would be treated independently. In addition we assume that all
NS's in a population will have the same deformability parameter $\lambdans$, and therefore
its cumulative measurement is possible.
% 
Using Bayes' theorem, the measured probability distribution function for $\lambdans$
from $N$ observations is
\begin{equation}\label{eq:lambdaMultiple}
 p(\lambdans | d_1, d_2, \cdots, d_N; K) = p(\lambdans | K)^{1-N}\prod_{i=1}^N p(\lambdans | d_i, K),
\end{equation}
where $p(\lambdans|K)$ is the prior expectation for $\lambdans$, and we have assumed
that all of the $N$ observations are independent. A priori, we assume that no 
particular value of $\lambdans$ is more preferred over another, and $\lambdans\leq 4000$,
i.e.
\begin{equation}\label{eq:lprior}
 p(\lambdans | K) = \dfrac{1}{4000}\,\mathrm{Rect}\left(\frac{\lambdans-2000}{4000}\right)
\end{equation}
% 
The second set of terms in Eq.~\ref{eq:lambdaMultiple} of the form $p(\lambdans | d_i, K)$
is each the probability distribution for $\lambdans$ inferred {\it a posteriori} from
the \textit{i}-th observation independently. 
They are calculated by integrating
\begin{equation}
 p(\lambdans | d_i, K) = \int \D \vec{\theta}\, p(\vec{\theta}, \lambdans | d_i, K),
\end{equation}
where $p(\vec{\theta}, \lambdans | d_n, K)$ is the inferred joint probability 
distribution of all source parameters $\vec{\theta}\cup\{\lambdans\}$ for the 
$i$-th event, as described in Sec.~\ref{s2:bayesian}.

Our goal is to understand the improvement in our measurement of $\lambdans$
with the number of recorded events. To do so, we simulate populations of $N$
observations each, and quantify what we learn from each successive observation 
using Eq.~\ref{eq:lambdaMultiple}. This allows us to determine how
rapidly does our median estimate for $\lambdans$ converges to the true value,
and how broad our uncertainty intervals for the same are, as $N$ increases.
In order to generate each population, we first fix
the NS properties: (i) NS mass $\mns=1.35M_\odot$, (ii) NS spin $\chins=0$
and (iii) NS tidal deformability $\lambdans=$ some value that is chosen from
$\{500,800,1000,1500,2000\}$. Next, we successively generate $N$ 
independent events, uniformly sampling the remaining
source parameters from the following ranges: (i) mass-ratio $q\in[2,5]$, (ii)
BH spin $\chibh\in[0, 0.75]$, (iii) effective distance from within a 
spherical volume whose radius is given by the lowest detectable
SNR (which we take as $\rho=10$). We remind ourselves that
once source parameters are fixed, there is a one-to-one mapping between
effective distance and SNR. Finally, in order to mitigate computational
cost, we make an additional simplification.
We perform full MCMC parameter estimation for a set of systems with parameters
laid on the regular hypercubic grid (henceforth ``G'') that spans
$q=\{2,3,4,5\}\times\chibh=\{-0.5,0,0.5,0.75\}\times\lambdans=\{500,800,1000,1500,2000\}\times\rho=\{10,20,30,50,70\}$.
The final population is given by substituting all of the $N$ events, generated
as above, with their respective nearest neighbours on the grid G.

In Fig.~\ref{fig:TT_Lambda_vs_N_L800_CI90_0} we show results for a population
of $N=50$ events with $\lambdans=800$. In the left panel, we show the 
probability distributions for $\lambdans$ inferred from $N$ events with $N$
ranging from $1-50$. We also mark the $90\%$ confidence interval on $\lambdans$
as dark grey vertical lines corresponding to each cumulative distribution.
Distributions are normalized to unit area. We see that the information coming
from the first few observations does not limit $\lambdans$ much more than our
prior from Eq.~\ref{eq:lprior} does. However, the median of the distribution 
tracks the true value of $\lambdans$ fast, reaching close within as few as $4$
observations.
% 
In the right panel, we show our measurement uncertainty $\delta\Lambda/\Lambda$
as a function of $N$, where we take $\delta\Lambda$ to be the width of the 
$90\%$ confidence interval on $\lambdans$ after $N$ observations.
This is given by the green filled region (left y-axis).
The green filled circles show the median measured value of $\lambdans$
as a function of $N$. The dashed line at y value $=1$ marks zero error, i.e.
$\delta\Lambda/\Lambda=0$. The dotted and dashed lines around this denote
$\pm 25\%$ and $\pm 50\%$ relative errors, respectively. Independently, the
light-grey filled shown the absolute measurement uncertainty $\delta\Lambda$
on the right y-axis , as a function of $N$. A horizontal grey dashed line
marks the true value of $\lambdans$. Contours of $1/\sqrt{N}$ are shown as 
dotted light grey curves to guide the eye while studying the dependence of
$\delta\Lambda$ on $N$.
%
